---
title: "Regression Project Markdown"
author: "Kathryn Saunders & Keegan Chatham"
date: "2025-02-24"
output: pdf_document
---

```{r setup, include=FALSE}
library(tidyverse)
library(knitr)
library(tinytex)
library(readxl)
library(ggplot2)
library(car)

knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(warning = FALSE)
knitr::opts_chunk$set(out.width = "60%",fig.align="center")
```

migration data
https://www.pewresearch.org/dataset/dataset-religious-composition-of-the-worlds-migrants-1990-2020/

gdp data
https://data.worldbank.org/indicator/NY.GDP.MKTP.CD

population data
https://data.worldbank.org/indicator/SP.POP.TOTL

## Assorted Wrangling (everything done before the project proposal)

```{r}

#read main data
migration <- read_excel("~/RStudio/PROJECT/Incoming_and_Outgoing_Migrant_Counts.xlsx")
view(migration)

#fix column issues
colnames(migration) <- c("direction", "year", "country", "religion", "count",
                         "percent", "region", "level", "countrycode")

migration2 <- migration[-1,]

#fix variable types
migration3 <- migration2 %>%
  mutate(direction = as.factor(direction),
         year=as.numeric(year),
         country=as.factor(country),
         religion=as.factor(religion),
         count=as.numeric(count),
         percent=as.integer(percent),
         region=as.factor(region),
         level=as.numeric(level),
         countrycode=as.numeric(countrycode))

#NOTE: Everywhere that count says 10,000 used to be < 10,000
migration3["count"][is.na(migration3["count"])] <- 10000



# IF IT WORKS WITH THESE COMMENTED OUT THEN DELETE THEM - DONE LATER
# gdp_simple2 <- read_excel("~/RStudio/PROJECT/GDPbyCountry_1990-2020.xls")
# 
# country_continents <- read_csv("~/RStudio/PROJECT/Countries-Continents.csv")
# population <- read_excel("~/RStudio/PROJECT/population2.xls")
# 
# cont_pop <- country_continents %>%
#   left_join(population, by= c("Country"="Country Name"))
# 
# 


#split into migration in and migration out then remerge for net migration
country_migration <- migration3 %>%
  filter(level==1) 

out_migration <- country_migration %>%
  filter(direction == "Outgoing") %>%
  rename(percent_out = percent, count_out = count)

out_migration2 <- out_migration[,-1]

in_migration <- country_migration %>%
  filter(direction == "Incoming") %>%
  rename(percent_in = percent, count_in = count)

in_migration2 <- in_migration[,-1]

net_migration <- in_migration2 %>%
  left_join(out_migration2, by=c("year", "country", "religion", "region", "level", "countrycode")) %>%
  mutate(net_migration = count_in - count_out)



#random sample from net_migration
set.seed(1234)
rows <- sample(x=1:nrow(net_migration),
               size=8000,
               replace=FALSE)

random_migration <- net_migration[rows,]


#select only majority religion
net_religion <- random_migration %>%
  filter(percent_in != 100) %>%
  group_by(country) %>%
  arrange(desc(percent_in)) %>%
  summarize(maj_religion=max(percent_in))

#add back to main dataframe
religion_migration <- random_migration %>%
  left_join(net_religion, by="country")

#select only the rows with the majority religion
religion_mig2 <- religion_migration %>%
  filter(percent_in==maj_religion)

#get rid of useless columns
religion_mig3 <- religion_mig2[,-c(7,8,12)]


#HAVE TO MERGE GDP AND POPULATION
  

#read gdp file
gdp <- read_excel("~/RStudio/PROJECT/GDPbyCountry_1990-2020.xls")

#switch rows and columns
gdp_long <- gdp %>%
  pivot_longer(cols = -1, names_to = "years", values_to = "gdp") %>%
  mutate(years=as.numeric(years))

#add to main data
religion_mig4 <- religion_mig3 %>%
  left_join(gdp_long, by=c("country" = "Country Name", "year" = "years")) %>%
  na.omit()

#read total population file (not migration)
population <- read_excel("~/RStudio/PROJECT/population2.xls")

#get rid of useless columns
pop2 <- population[,-c(2:4)]

#switch rows and columns
pop_long <- pop2 %>%
  pivot_longer(cols = -1, names_to = "years", values_to = "population") %>%
  mutate(years=as.numeric(years))

#add to main data
religion_mig5 <- religion_mig4 %>%
  left_join(pop_long, by=c("country" = "Country Name", "year" = "years")) %>%
  na.omit()

```


```{r, making it linear}
numeric_only <- religion_mig5[,-c(2,3,6)]
plot(numeric_only)

linear06 <- numeric_only %>% 
  transmute(year, loggdp = log(gdp), logcountin = log(count_in), 
            logperin = log(percent_in), logcountout = log(count_out), 
            logperout = log(percent_out), net_migration, 
            logpop = log(population))

#it looks so much better
plot(linear06)


#log model is more significant in everything than the non-log model
non_linearmodel01 <- lm(gdp~., data=numeric_only)
summary(non_linearmodel01)

linearmodel01 <- lm(loggdp~., data=linear06)
summary(linearmodel01)

vif(linearmodel01)

#adding the qualitative variables back in
linear07 <- religion_mig5 %>% 
  transmute(year, country, religion, region, loggdp = log(gdp), logcountin = log(count_in), 
            logperin = log(percent_in), logcountout = log(count_out), 
            logperout = log(percent_out), net_migration, 
            logpop = log(population))


```


```{r, collinearity problems take 2}
# removing count_out because of high vif score and low significance

categorical_tests <- religion_mig5[, -c(1,4,5,7,8,9,11)]

catmod01 <- lm(gdp~., data=categorical_tests)
summary(catmod01)

#definitely can not have both religion and country
#vif(catmod01)
#alias(catmod01)

#removed country
linear08 <- religion_mig5 %>% 
  transmute(year, religion, region, loggdp = log(gdp), logcountin = log(count_in), 
            logperin = log(percent_in), logcountout = log(count_out), 
            logperout = log(percent_out), net_migration, 
            logpop = log(population))

mod02 <- lm(loggdp~., data=linear08)
summary(mod02)


#the religiously unaffiliated countries are just China, Japan, and Sweden
#only unaffiliated religion is significant so I'm making an indicator variable

linear09 <- religion_mig5 %>%
    transmute(year, country, unaffiliated = ifelse(religion == "Unaffiliated", 1, 0), 
              region, loggdp = log(gdp), logcountin = log(count_in), 
              logperin = log(percent_in), logcountout = log(count_out), 
             logperout = log(percent_out), net_migration, 
             logpop = log(population))
  

#country is still ruining religion and region so lets see which is more accurate
mod03 <- lm(loggdp~., data=linear09)
summary(mod03)  

#NOTE: everything except net_migration is highly significant, RSE = 0.8246, R^2 = 0.8904
#has religion and region
no_country <- linear09 %>%
  select(-country)

mod04 <- lm(loggdp~., data=no_country)
summary(mod04) 

#better without useless net_mig - RSE = 0.8244 , R^2 = 0.8905 
no_country2 <- no_country[,-9]
mod04_5 <- lm(loggdp~., data=no_country2)
summary(mod04_5)

#really only some of the countries and logpop are significant, RSE = 0.3586, R^2 = 0.979
# has country but not religion or region
# HAS OVERFITTING ISSUES
yes_country <- linear09 %>%
  select(-unaffiliated, -region)

mod05 <- lm(loggdp~., data=yes_country)
summary(mod05) 


```


# NOTE - WORRIED ABOUT OVERFITTING THE SECOND MODEL - TRY SOME SHIT

```{r, no country RMSE - does work better option}
#randomly select rows from sample data
set.seed(123)
no_rows <- sample(x=1:nrow(no_country),size=floor(0.75*nrow(no_country)),replace=FALSE)


#split sample into training and testing
no_training <- no_country[no_rows,]
no_testing <- no_country[-no_rows,]

#check
mean(no_training$loggdp)
mean(no_testing$loggdp)

#compute RMSE for null model - 2.679811
no_testing %>%
  summarize(RMSE=sqrt(mean((loggdp-22.99685)^2)))


#estimate simple linear regression model
mod06 <- lm(loggdp~., data=no_training)

#estimate accuracy of model - 0.7480504
no_testing %>%
  mutate(pred_gdp=predict(mod06,newdata=no_testing)) %>%
  summarize(RMSE=sqrt(mean((loggdp-pred_gdp)^2)))

```


```{r, yes country rmse - doesn't work}
#randomly select rows from sample data
set.seed(123)
yes_rows <- sample(x=1:nrow(yes_country),size=floor(0.75*nrow(yes_country)),replace=FALSE)


#split sample into training and testing
yes_training <- yes_country[yes_rows,]
yes_testing <- yes_country[-yes_rows,]

#check
mean(yes_training$loggdp)
mean(yes_testing$loggdp)

#compute RMSE for null model - 2.679811
yes_testing %>%
  summarize(RMSE=sqrt(mean((loggdp-22.99685)^2)))


#estimate simple linear regression model
mod07 <- lm(loggdp~., data=yes_training)

#estimate accuracy of model
yes_testing %>%
  mutate(pred_gdp_yes=predict(mod07,newdata=yes_testing)) %>%
  summarize(RMSE=sqrt(mean((loggdp-pred_gdp_yes)^2)))

## last bit won't run because the testing data is intriducing new countries it hasn't seen before

```

## conclusion: use region and religion, don't use country


# basic min max stuff for methods section

```{r, for methods section table}

methods01 <- no_country2[, c(4,5,7,9)]

min(methods01$loggdp)
max(methods01$loggdp)
median(methods01$loggdp)

min(methods01$logpop)
max(methods01$logpop)
median(methods01$logpop)

min(methods01$logcountin)
max(methods01$logcountin)
median(methods01$logcountin)

min(methods01$logcountout)
max(methods01$logcountout)
median(methods01$logcountout)

```

# graphics for exploratory section

```{r, exploratory graphics}
#The first graphic must display the distribution of the response variable.
graph1 <- ggplot(no_country2, aes(loggdp)) + geom_histogram(color="black",fill="#CF887C") + 
  theme_bw() +
  labs(x = "The Log of GDP (Current USD)",
       y = "Count",
       title = "GDP Distribution",
       subtitle = "Figure 1")
graph1

ggsave("Stat3400_ProjGraphic1.jpg", plot = graph1, width = 8, height = 6, units = "in", dpi = 300)

#The second graphic must display the distribution of one important predictor.
graph2 <- ggplot(no_country2, aes(logcountin)) + geom_histogram(color="black",fill="#CF887C", bins=20) + theme_bw() +
  labs(x = "The Log of the Number of Incoming Migrants",
       y = "Count",
       title = "Distribution of the Number of Immigrants",
       subtitle = "Figure 2")
graph2

ggsave("Stat3400_ProjGraphic2.jpg", plot = graph2, width = 8, height = 6, units = "in", dpi = 300)

#The third graphic must display the association between the response and predictor.
graph3 <- ggplot(no_country2, aes(x=logcountin, y=loggdp)) + geom_point() + 
  geom_jitter(width=0.1,height=0,alpha=0.5) + theme_bw() +
  geom_smooth(method="lm",formula=y~x, se=FALSE, color="#E75A50") +
  labs(x = "The Log of the Number of Incoming Migrants",
       y = "The Log of GDP (Current USD)",
       title = "Immigration Patterns in GDP",
       subtitle = "Figure 3")

ggsave("Stat3400_ProjGraphic3.jpg", plot = graph3, width = 8, height = 6, units = "in", dpi = 300)


#The fourth graphic can display any distribution or association you find interesting.
graph4 <- ggplot(no_country2, aes(x=region, y=loggdp)) + geom_boxplot(color="black",fill="#CF887C") + 
  theme_bw() + theme(axis.text.x = element_text(angle = 20, hjust = 1, vjust = 1)) +
  labs(x = "Region",
       y = "The Log of GDP (Current USD)",
       title = "GDP by World Region",
       subtitle = "Figure 4")
graph4

ggsave("Stat3400_ProjGraphic4.jpg", plot = graph4, width = 8, height = 6, units = "in", dpi = 300)


regions <- no_country2 %>%
  filter(region=="Middle East-North Africa")
median(regions$loggdp)


#AP 21.99
#E 23.82
#LAC 22.29
#MENA 24.609
#NA 25.204
#SSA 22.51
```





# PREDICITVE ANALYSIS SECTION - USE NO_COUNTRY2 AS THE STARTING DATA SET BC I DONT WANT TO FIGURE THAT OUT AGAIN

```{r, diagnostic plot}
#fit linear regression model
diagnostic_model <- lm(loggdp~.,data=no_country2)

#display first diagnostic plot
jpeg('diagnostic.jpg', width = 8, height = 6, units = "in", res=300)
plot(diagnostic_model,which=1)



```

```{r, stepwise variable selection}
glimpse(no_country2)

library(olsrr)

step_model01 <- lm(loggdp~., data=no_country2)

forward_model01 <- ols_step_forward_adj_r2(step_model01)

forward_model01$metrics

backward_model01 <- ols_step_backward_adj_r2(step_model01)

backward_model01$metrics

#all of the 8 variables both ways, what a surprise.

#construct data frame
forward_results <- data.frame(vars=seq(0,8,1),
                              adjR=c(0,forward_model01$metrics$adj_r2))

#plot forward stepwise results
graph6 <- ggplot(data=forward_results,
       mapping=aes(x=vars,y=adjR)) +
  geom_line(color="#E75A50", size=1) +
  geom_hline(yintercept=max(forward_results$adjR),
             color="black",linetype="dashed") +
  labs(title="Forward Stepwise Selection",
       x="Number of Predictors",
       y="Adjusted R-Squared",
       subtitle = "Figure 6") +
  scale_x_continuous(limits=c(0,10),breaks=seq(0,10,1)) +
  theme_bw()

graph6

ggsave("Stat3400_ProjGraphic6.jpg", plot = graph6, width = 8, height = 6, units = "in", dpi = 300)


```

```{r, cross validation - GRAPHIC?}
library(boot)

#null model
null_rmse <- sqrt(mean((no_country2$loggdp-22.95574)^2))
null_rmse

#LOOCV cross-validation
loocv_model <- glm(loggdp~., data=no_country2, family="gaussian")

loocv <- cv.glm(data=no_country2,glmfit=loocv_model,K=nrow(no_country2))

mse <- loocv$delta[1]

sqrt(mse)

#5-fold CV
set.seed(303)
five_cv <- cv.glm(data=no_country2,glmfit=loocv_model,K=5)

#extract RMSE
sqrt(five_cv$delta[1])

#null RMSE is 2.487708
#LOOCV RMSE is 0.85437
#5-fold CV RMSE is 0.8331947 - lower so might use this lol

#comparing RMSE or R-squared of different model options - null, best linear model with no log, best with log
```

```{r, cv part 2}
complinear <- no_country2 %>%
  transmute(year, unaffiliated, region, loggdp, countin = exp(logcountin),
            perin = exp(logperin), countout = exp(logcountout), 
            perout = exp(logperout), pop = exp(logpop))

compnull_rmse <- sqrt(mean((complinear$loggdp-22.95574)^2))
compnull_rmse

#okay same null - good

comp_model <- glm(loggdp~., data=complinear, family="gaussian")
summary(comp_model)


#5-fold CV
set.seed(303)
compfive_cv <- cv.glm(data=complinear,glmfit=comp_model,K=5)

#extract RMSE
sqrt(compfive_cv$delta[1])

#1.82688 - RMSE of 5-fold CV with non-logged variables

```

```{r, cv graphic}

cv_graph <- data.frame(
  cv = c("Null Model", "Non-Linearized Model", "Linearized Model"),
  rmse = c(2.487708, 1.82688, 0.8331947))

graph7 <- ggplot(cv_graph, aes(x=cv, y=rmse)) + geom_histogram(stat = "identity",fill="#CF887C") +
  theme_bw() + labs(
    title = "5-Fold Cross Validation Error",
    subtitle = "Figure 7",
    x = "Model",
    y = "Root Mean Squared Error")

graph7

ggsave("Stat3400_ProjGraphic7.jpg", plot = graph7, width = 8, height = 6, units = "in", dpi = 300)


```


## fourth interesting thing
```{r}
step_model01 <- lm(loggdp~., data=no_country2)

no_country3 <- no_country2 %>%
  mutate(predicted = predict(step_model01))

graph8 <- ggplot(no_country3, aes(x = predicted, y = loggdp)) +
  geom_point(position="jitter", alpha=0.5) + geom_jitter(width=0.1,height=0,alpha=0.5) +
  geom_abline(slope = 1, intercept = 0, color = "#E75A50") +
  labs(title = "Actual vs Predicted Values of the Log of GDP",
       subtitle = "Figure 8",
       x = "Predicted Values",
       y = "Actual Values") +
  theme_bw()

graph8

ggsave("Stat3400_ProjGraphic8.jpg", plot = graph8, width = 8, height = 6, units = "in", dpi = 300)

```





# DISCUSSION TABLE
```{r}
# Adjusted R^2 - 0 versus 0.891
# RSS - 2.491 versus 0.8244
# ANOVA P-VALUE - NA versus 2.2e-16

null_model <- lm(loggdp~1, data=no_country2)
summary(null_model)

discussion_model <- lm(loggdp~., data=no_country2)
summary(discussion_model)


# CROSS VALIDATED ERROR (from above) - 2.488 versus 0.833


```









## ignore these two sections

```{r, multicollinearity - OLD IGNORE, include=FALSE}
#very much a mess
model0 <-lm(gdp~., data=religion_mig5)
summary(model0)

#trying to figure out whether percent in/out or count in/out is better (multicollinearity)
numeric_only <- religion_mig5[,-c(2,3,6)]

model01 <- lm(gdp~., data=numeric_only)
summary(model01)

#count_in is better than percent_in so lets just use that
#having both count_in and count_out is giving multicollinearity errors with net migration so I'm removing count_out

religion_mig6 <- religion_mig5[,-c(2,5, 7,8)]
model03 <- lm(gdp~., data=religion_mig6)
summary(model03)
vif(model03)
alias(model03)

#update having count_in and count_out are also having mutlicollinearity issues, going down to just count_in
religion_mig7 <- religion_mig5[,-c(2,5,7,8,9)]
model04 <- lm(gdp~., data=religion_mig7)
summary(model04)
vif(model04)
alias(model04)
#THE VIFs ARE FINALLY BELOW 5


#nevermind putting country back in brought the alias error back
religion_mig8 <-religion_mig5[,-c(2,5,7,8,9)]
model05 <- lm(gdp~., data=religion_mig8)
summary(model05)
vif(model05)
alias(model05)


```

```{r, ignore this, include=FALSE}
numeric_only2 <- religion_mig8[,-c(2,4)]
plot(numeric_only2)

log_numeric <- numeric_only2 %>%
  mutate(loggdp = log(gdp), logpop = log(population), logcount = log(count_in))

plot(log_numeric)

```







